Analysis of $\piup$-$\piup$ lattice QCD \\ scattering data

### 2015-03-18 Paper.tex
Improve small things
Write a tad more

Monte Carlo methods can be used to simulate quantum systems. Key pieces are
Feynman's path integral formalism and Euclidean spacetime where the metric
tensor $\tens\eta$ has signature $\pm 4$ and the weight factor in the path
integral is $\exp(-S_\text{cl})$ instead of $\exp(\iup S_\text{cl})$ where
$S_\text{cl}$ is the classical action
\parencite[Section~2]{Creutz/Statistical_Approach_QM}. A Monte Carlo algorithm
like the Metropolis algorithm can then be used to generate configurations which
are already distributed by the weight of the classical action. Therefore,
expectational values of observables are simply calculated and averaged over all
available configurations \parencite[(3.7)]{Creutz/Statistical_Approach_QM}.

The group of Andreas Kell, Martin Efferz and Simon Blanke have used this for
the harmonic oscillator in their Computational Physics project this year. I
also did this in my bachelor's thesis: \parencite{Ueding/Bachelorarbeit}.

@@

As with the harmonic oscillator on the lattice, expectation values of
observables can be computed for each configuration of the ensembles and then
averaged (without weight) over all ensembles yielding an estimate for the
observable. Since this uses a field theory, particles of interested can be
created at the beginning of time and annihilated at the end of the temporal
lattice extent. This way, specific particle configurations (like
$\piup$–$\piup$ scattering) can be examined without needing to generate new
configurations.

@@

The data that I was given is organized in different ensembles (like A30.32,
A100.24). For each ensemble, multiple configurations were simulated. The
output of each configuration is one two-point correlation function $C_\piup(t)$
and three four-point correlation functions $C_{\piup\piup}^{(i)}(t)$. Those
three different contractions are combined into a single
$\piup\piup$-correlation function:
\begin{equation}
    C_{\piup\piup}(t) = C_{\piup\piup}^{(1)}(t) + C_{\piup\piup}^{(2)}(t)
    - 2 C_{\piup\piup}^{(3)}(t).
\end{equation}

@@

In both space and time, the lattice used in the computation has periodic
boundary conditions. The spatial lattice extent is called $L$, the temporal one
$T$. This leads to a symmetry where the time slices $t$ and $T-t$ closely
related. The correlation functions $C(t)$ with $t = |t_1 - t_2|$ are symmetric
in $t_1$ and $t_2$, such that only the data in the interval $[0, T/2]$ carries
independent information. Before starting any other calculations with the data,
it is folded in half and averaged. The data then looks like in
Figure~\ref{fig:folded} which is shown later on.

@@

The pion masses are contained in the correlation functions in the form of an
exponential decay constant. Since time is periodic in this simulation, the
correlation function is also periodic with the temporal lattice extent $T$.
Therefore, the expectation is not a simple exponential decay, $\exp(-\lambda
t)$, but a cosh-like function. The four-point functions also have a constant
contribution due to the finite lattice extent. For both two and four point
functions, the functions that I fitted to the folded data with parameters $\vec
\lambda$ are:
\begin{equation}
    \label{eq:fit2}
    m_\piup(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])}
\end{equation}
and
\begin{equation}
    \label{eq:fit4}
    m_{\piup\piup}(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])} + \lambda_3.
\end{equation}

@@

For the correlated fit, a new likelihood function is needed. I chose to keep
the least squared likelihood function but incorporate the correlation into the
$\chi^2$. First, a correlation matrix $\tens C$ is needed, which is computed
from the $R$ bootstrap samples: \parencite[Section~2]{Michael:1994sz}
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{r=1}^R
    [x_{ir} - \bar x_{iR}] [x_{jr} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

In the curve fitting process, a function like
\texttt{scipy.optimize.curve\_fit} tries to find the parameters $\lambda$ such
that $\chi^2$ becomes minimal. In my implementation, I used
\texttt{scipy.optimize.leastsq} which tries to minimize the squared norm of a
vector valued function. With a Cholesky decomposition, this can be done like
so: Define the square bracket in Equation~\eqref{eq:chisq} to be the residual
vector $\vec r$. Then the matrix multiplication can be written as
\begin{equation}
    \chi^2 = \vec r^\text T \tens C^{-1} \vec r.
\end{equation}
The Cholesky decomposition gives me $\tens C^{-1} = \tens U^\dagger \tens U$,
where $\tens U$ is a upper triangle matrix. Then I can write
\begin{equation}
    \chi^2 = [\tens U \vec r]^\dagger [\tens U \vec r]
    = \| \tens U \vec r \|^2
\end{equation}
where $\tens U \vec r$ is a vector that can be fed into
\texttt{scipy.optimize.leastsq}.

@@

Equation~\eqref{eq:luescher} can be motivated like so: Let $\hat H = \hat H_0 +
\hat V$ be the Hamiltonian for two identical particles where $\hat H_0$ is the
free Hamiltonian and $\hat V$ a spherically symmetric potential with a limited
range (or faster than $1/\sqrt r$ decay). Then the probability that the two
particles are in interaction range scales like $L^{-3}$ as this is the inverse
volume which the two particles occupy. Then the most prominent finite size
effect scales to the third power, all other effects can only have higher
powers.

In the first order of the potential $V$ the ground state (all momenta are zero)
will have an energy shift proportional to $L^{-3}$
\parencite[(2.24)]{luescher/volume_dependence}:
\[
    \Deltaup E = \frac{1}{2L^3} \hat V(\vec 0, \vec 0) + \mathcal O(V^2).
\]
The scattering amplitude $T$ as well as the S-wave scattering length $a_0$ can
be expanded in powers of $\hat V$ (Born series). Using first order term from
\parencite[(2.18)]{luescher/volume_dependence}, above equation can be rewritten
with the scattering length
\parencite[(2.25)]{luescher/volume_dependence}:
\[
    \Deltaup E = - \frac{4\piup a_0}{m L^3} + \mathcal O(V^2).
\]

To first order on $V$, the leading term scales with $L^{-3}$, like in
Equation~\eqref{eq:luescher}. To obtain more terms, a complete perturbation
expansion needs to be written out. That expansion will be both in terms of
$L^{-1}$ and $V$. \citeauthor{luescher/volume_dependence} then was able to
group those terms in powers of $L^{-1}$ and obtain
Equation~\eqref{eq:luescher}.

@@

>>>>

Monte Carlo methods can be used to simulate quantum systems. Key pieces are
Feynman's path integral formalism and Euclidean spacetime. In the latter, the metric
tensor $\tens\eta$ has signature $\pm 4$ and the weight factor in the path
integral is $\exp(-S_\text{cl})$ instead of $\exp(\iup S_\text{cl})$; 
$S_\text{cl}$ is the classical action
\parencite[Section~2]{Creutz/Statistical_Approach_QM}. A Monte Carlo algorithm
like the Metropolis algorithm can then be used to generate configurations which
are already distributed by the weight of the classical action. Therefore,
expectational values of observables are simply calculated and averaged over all
available configurations \parencite[(3.7)]{Creutz/Statistical_Approach_QM}.

The group of Andreas Kell, Martin Efferz and Simon Blanke have used this for
the harmonic oscillator in their Computational Physics project this year. I
also did this in my bachelor's thesis last year \parencite{Ueding/Bachelorarbeit}.

@@

As with the harmonic oscillator on the lattice, expectation values of
observables can be computed for each configuration of the ensembles and then
averaged (without weight) over all ensembles yielding an estimate for the
observable. Since this uses a field theory, particles of interest can be
created at the beginning of time and annihilated laster on. This way, specific
particle configurations (like $\piup$–$\piup$ scattering) can be examined
without needing to generate new configurations.

@@

The data that I was given is organized in different ensembles (like A30.32,
A100.24) as mentioned in the introduction. For each ensemble, multiple
configurations were simulated. The output of each configuration is one
two-point correlation function $C_\piup(t)$ and three four-point correlation
functions $C_{\piup\piup}^{(i)}(t)$. Those three different contractions are
combined into a single $\piup\piup$-correlation function:
\begin{equation}
    C_{\piup\piup}(t) = C_{\piup\piup}^{(1)}(t) + C_{\piup\piup}^{(2)}(t)
    - 2 C_{\piup\piup}^{(3)}(t).
\end{equation}

@@

In both space and time, the lattice used in the computation has periodic
boundary conditions. The spatial lattice extent is called $L$, the temporal one
$T$. This leads to a symmetry where the time slices $t$ and $T-t$ closely
related. The correlation functions $C(t)$ with $t = |t_1 - t_2|$ are symmetric
in $t_1$ and $t_2$, such that only the data in the interval $[0, T/2]$ carries
independent information. Before starting any other calculations with the data,
it is folded in half and averaged. The data then looks like in
Figure~\ref{fig:folded} which appears laster on.

@@

The pion masses are contained in the correlation functions in the form of an
exponential decay constant. Since time is periodic in this simulation, the
correlation function is also periodic with the temporal lattice extent $T$.
Therefore, the expectation is not a simple exponential decay, $\exp(-\lambda
t)$, but a cosh-like function. The four-point functions also have a constant
contribution due to the finite lattice extent. For both two and four point
functions, the functions that I fitted to the folded data with parameters $\vec
\lambda$ are:
\begin{equation}
    \label{eq:fit2}
    m_\piup(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])}
\end{equation}
and
\begin{equation}
    \label{eq:fit4}
    m_{\piup\piup}(t, \lambda) = \lambda_1 \sbr{\exp(-\lambda_2 t) + \exp(-\lambda_2
    [T - t])} + \lambda_3.
\end{equation}
The parameter $\lambda_3$ is a lattice artifact that will vanish for $L \to
\infty$.

@@

For the correlated fit, a new likelihood function is needed. I chose to keep
the falling exponential likelihood function but incorporate the correlation
into the $\chi^2$. First, a correlation matrix $\tens C$ is needed, which is
computed from the $R$ bootstrap samples: \parencite[Section~2]{Michael:1994sz}
\begin{equation}
    \label{eq:correlation_matrix}
    C_{ij} := \frac{1}{R[R-1]} \sum_{r=1}^R
    [x_{ir} - \bar x_{iR}] [x_{jr} - \bar x_{jR}]
    \eqnsep
    \bar x_{iR} := \frac 1R \sum_{r=1}^R x_{ir}.
\end{equation}

@@

In the curve fitting process, a function like
\texttt{scipy.optimize.curve\_fit} tries to find the parameters $\vec\lambda$
such that $\chi^2$ becomes minimal. In my implementation, I used the function
\texttt{scipy.optimize.leastsq} which tries to minimize the squared norm of a
vector valued function. With a Cholesky decomposition, this can be done like
so: Define the square bracket in Equation~\eqref{eq:chisq} to be the residual
vector $\vec r(\vec \lambda)$. Then the matrix multiplication can be written as
\begin{equation}
    \chi^2 = \vec r^\text T \tens C^{-1} \vec r.
\end{equation}
The Cholesky decomposition gives me $\tens C^{-1} = \tens U^\dagger \tens U$,
where $\tens U$ is a upper triangle matrix. Then I can write
\begin{equation}
    \chi^2 = [\tens U \vec r]^\dagger [\tens U \vec r]
    = \| \tens U \vec r \|^2
\end{equation}
where $\tens U \vec r$ is a vector that can be fed into
\texttt{scipy.optimize.leastsq}.
\nocite{SciPy}

@@

Equation~\eqref{eq:luescher} can be motivated like so: Let $\hat H = \hat H_0 +
\hat V$ be the Hamiltonian for two identical particles where $\hat H_0$ is the
free Hamiltonian and $\hat V$ a spherically symmetric potential with a limited
range (or faster than $1/\sqrt r$ decay). Then the probability that the two
particles are in interaction range scales like $L^{-3}$ as this is the inverse
volume which the two particles occupy. The most prominent finite size effect
scales to the third power, all other effects can only have higher powers of
$L^{-1}$.

In the first order of the potential $V$, the ground state (all momenta are zero)
will have an energy shift proportional to $L^{-3}$
\parencite[(2.24)]{luescher/volume_dependence}:
\[
    \Deltaup E = \frac{1}{2L^3} \hat V(\vec 0, \vec 0) + \mathcal O(V^2).
\]
The scattering amplitude $T$ as well as the S-wave scattering length $a_0$ can
be expanded in powers of $\hat V$ (Born series). Using first order term from
\parencite[(2.18)]{luescher/volume_dependence}, above equation can be rewritten
with the scattering length
\parencite[(2.25)]{luescher/volume_dependence}:
\[
    \Deltaup E = - \frac{4\piup a_0}{m L^3} + \mathcal O(V^2).
\]

To first order on the potential $V$, the leading term scales with $L^{-3}$,
like in Equation~\eqref{eq:luescher}. To obtain more terms, a complete
perturbation expansion needs to be written out. That expansion will be both in
terms of $L^{-1}$ and $V$. \citeauthor{luescher/volume_dependence} then was
able to group those terms in powers of $L^{-1}$ and obtain
Equation~\eqref{eq:luescher}.

@@

The range of fitting was set manually by looking at the plateau of the
effective masses. This introduces some statistical errors which cannot be
quantified with the current machinery. A way to get beyond this limitation is
to run the whole analysis with all possible fit ranges. The different result
are then weighted by the $p$-values of the respective fit.

\clearpage
